---
title: "Statistical Analysis, MSCA 31007, Lecture 3"
author: "Hope Foster-Reyes"
date: "October 28, 2016"
output:
  pdf_document:
    fig_width: 6.5
    fig_height: 4.5
geometry: margin=0.75in
---

# Simulation of 4 Linear Models

Understand common assumptions behind the linear model.

Set the values of slope (a) and intercept (b) parameters. Set the sample length to 1,000.

```{r setup}
# Slope and Intercept
a <- 0.8; b <- 0.1
sample.n <- 1000

# Seeds
seed1 <- 111
seed2 <- 1112131415
```

## Model 1.

Simulate and plot Model 1:

Input variable $X \sim Norm(\mu= 3, \sigma=2.5)$; model residuals $Eps \sim Norm(\mu = 0, \sigma = 1.5)$

```{r m1, fig.width=6.5, fig.height=4.5}
# Simulate the model using rnorm()
set.seed(seed1)
model1.x <- rnorm(sample.n, mean = 3, sd = 2.5)
set.seed(seed2)
model1.eps <- rnorm(sample.n, mean = 0, sd = 1.5)

model1.y <- (a * model1.x + b) + model1.eps
model1 <- data.frame(Y=model1.y, X=model1.x, Eps=model1.eps)

# Inspect results
title <- "Model 1"
head(model1)
plot(model1$X, model1$Y, main = title)
plot(model1$Eps, type = "l", main = paste(title, "Residuals"))
```

## Model 2.

Simulate and plot Model 2:

Input variable $X \sim Norm(\mu= 3, \sigma=2.5)$; model residuals $Eps \sim Unif(min=-4.33, max=4.33)$

```{r m2}
set.seed(seed2)
model2.eps <- runif(sample.n, min = -4.33, max = 4.33)

model2.y <- (a * model1.x + b) + model2.eps
model2 <- data.frame(Y=model2.y, X=model1.x, Eps=model2.eps)

# Inspect results
title <- "Model 2"
head(model2)
plot(model2$X, model2$Y, main = title)
plot(model2$Eps, type = "l", main = paste(title, "Residuals"))
```

## Model 3.

Simulate and plot Model 3:

Input variable $X \sim Norm(\mu= 3, \sigma=2.5)$; model residuals $Eps \sim Cauchy*(location=0, scale=3)$

```{r m3}
set.seed(seed2)
model3.eps <- rcauchy(sample.n, location = 0, scale = 0.3)

model3.y <- (a * model1.x + b) + model3.eps
model3 <- data.frame(Y=model3.y, X=model1.x, Eps=model3.eps)

# Inspect results
title <- "Model 3"
head(model3)
plot(model3$X, model3$Y, main = title)
plot(model3$Eps, type = "l", main = paste(title, "Residuals"))
```